Death By Dropdown?

### 2025-04-08 conference-materials/2025_shiny-conf/slides.qmd
Refines speaker notes for clarity and conciseness
- Updates speaker notes in conference slides to improve clarity and engagement.
- Simplifies language by removing unnecessary phrases and quotes, focusing on key points and psychological principles.
- Enhances presentation flow and audience comprehension by streamlining content.

- Even though SoundPulse is a composite client...this is representative of many dashboards we've seen across the market research industry
- Point out key issues:
  * "Filter overload before seeing ANY insights"
  * "Multiple tabs with no clear starting point"
  * "Complete cognitive overload"
- "This is why dashboards get abandoned - nobody wants this experience"
- "Throughout the presentation, we'll show you the transformation process and the principles that guided our decisions"
:::

@@

- "The 5 stages form an easy workflow: Notice → Interpret → Structure → Anticipate → Validate"
  * Click through
- "Notice is about identifying objective friction points - where users struggle with the interface"
- "Interpret goes deeper to understand what story or insight users truly need - moving from symptoms to diagnosis"
- "Structure focuses on organizing information in ways that align with how our brains process information"
- "Anticipate considers how cognitive biases might affect data interpretation"
- "Validate ensures users leave with clear takeaways and a positive final impression"
- "These stages are iterative - you'll often revisit earlier stages as you discover new insights"
- "This diagram will serve as our roadmap throughout the presentation"
:::

@@

- "BID integrates existing best practices from multiple disciplines"..."We are truly standing on the shoulders of giants here"
  * Click through example fields...UX, data storytelling, psych science, data/visual communication
- Key differentiator: "But what makes BID uniquely powerful is how it systematically connects psychological principles to specific coding decisions at each stage of development"
- "Unlike checklists that tell you what to do, BID guides how you think about your dashboard from psychological first principles"
- "The framework bridges theory and practice - showing not just what principles matter, but exactly when and how to apply them in your development workflow"
- "Even though there's a lot of theory, you don't need a psychology degree to apply these principles effectively"
- "To demonstrate...let's take a closer look at how we connected these stages to SoundPulse's dashboard."
:::

@@

- **Stage 1 (Notice):** Users were overwhelmed with options
  * This is related to 'cognitive load' - which is the mental effort needed to process information
  * When the brain gets overtaxed, comprehension drops
  * Hick's Law also shows decision time increases logarithmically with the number of choices 
  * So more options, means exponentially more confusion
- **Stage 2 (Interpret):** Taking this to the next step helped us interpret what the user needed: Increased Processing Fluency
  * Processing fluency is the ease with which information is processed - when something is easy to process, people find it more credible and trustworthy
  * We also identified the key story the executives needed to understand...rather than showing all possible data
- **Stage 3 (Structure):** Moving to the Structuring stage...applied the 'Principle of Proximity' - which is a fancy way of saying we started grouping related elements together. 
  * Dual-Processing Theory distinguishes between fast, intuitive thinking and slower, deliberative thinking - we designed for both
  * We created groups of primary and secondary metrics to create a natural flow through the data
- "Now we'll show you the actual code implementation of these principles"
:::

@@

- "For the later stages of the BID framework, we addressed more subtle psychological factors that affect how users interpret and act on data."
- **Stage 4 (Anticipate):** We noticed executives were making flawed comparisons without proper context.
  * This relates to the 'Anchoring Effect' - where people rely too heavily on the first piece of information they see
  * We also saw teams struggling to interpret the data in the same way, so we addressed this with something called a "framing toggle"...more on that in just a moment
- **Stage 5 (Validate):** Finally, we applied the 'Peak-End Rule' - the idea that people judge experiences mainly by their most intense moment and ending
  * We created summary cards highlighting key takeaways so users leave with a clear, memorable impression
:::

@@

- "This is a great example to illustrate the theory behind these concepts"
- "In this case we are showing the *Same exact data*...but users experience completely different psychological state"
- "Progress framing creates motivation and highlights achievement"
  - "Marketing preferred progress framing to build stakeholder confidence"
- "Gap framing creates urgency and focuses on what remains to be done"
  - "Product teams used gap framing to prioritize development"
- "This isn't just theory - a classic study in the Judgment and Decision Making literature presented identical medical treatment outcomes as either a 90% survival rate or a 10% mortality rate - people react completely differently to these framings"
- "This is why understanding psychological principles is so powerful for dashboard design"
- "Think about your own dashboards - are you framing data in a way that drives the right decisions?"
:::

@@

- "We're not just sharing a framework or theories - we're giving you tools to implement it immediately"
- "The `{bidux}` package directly implements the 5-stage BID framework we've discussed today"
- "Each function maps to a specific stage and guides your thinking through that part of the development process"
- "Phase 1 is available today on GitHub - it provides the conceptual foundation and workflow guidance"
- "Future phases will include LLM integration to suggest specific implementations based on your needs"
- "We're also developing a UI component library with pre-built elements that embody these principles"
- "Think of `{bidux}` as your behavioral science assistant for Shiny app development"
:::

@@

- "Let's look at what {bidux} offers today"
- "The concept dictionary provides implementation advice, not just theory"
- "You can search across psychology, UI/UX, and data visualization fields...and more!"
- "The workflow functions guide you through each BID stage"
- "Auto-suggestion of theories based on your problem description"
- "Component suggestions provide concrete UI implementation ideas"
- "The package helps connect psychological principles to specific suggestions and components"
- "While we don't have time to go through a complete workflow demonstration, the package documentation on GitHub provides detailed examples"
- When the link appears: "We're launching this at ShinyConf today - check out the repo for full documentation"
:::

@@

- "The results speak for themselves"
- "While our quantitative metrics are still being collected through new telemetry systems we're implementing, qualitative feedback has been overwhelmingly positive"
- "Users who previously abandoned the dashboard now rely on it daily"
- "Decisions that took days now happen in minutes"
- It's still early, but the emerging message from our clients is clear: this framework is changing how teams interact with their data
- When testimonial appears: "This transformation is what BID delivers"
- "Remember, this wasn't just about visual improvements - this was a fundamental shift in how our users...and developers...approached the dashboard"
:::

@@

- We'd love to hear about your dashboard challenges"
- "Don't forget to follow us on LinkedIn/GitHub for updates on {bidux} development"
- "Feel free to connect with us for any questions or deeper discussions about implementation details we couldn't cover today"
- Quick answers to likely questions:
  * "How do you measure success?" → "Start with user testing and session analytics; we're building telemetry tools to help quantify impact"
  * "Does this work for technical audiences?" → "Absolutely, technical users appreciate clarity too - in fact, they often have the highest cognitive load due to the complexity of their work"
  * "How long does implementation take?" → "Start small - even one principle makes a difference. A complete BID implementation can take 1-2 sprints depending on dashboard complexity"
  * "How does this apply to domains beyond music?" → "We've successfully applied these principles in healthcare, finance, and retail dashboards - the psychological principles are universal"
:::

>>>>

- Even though SoundPulse is a composite client...this dashboard from SoundPulse represents many we've seen across industries
- Notice the key issues: filter overload before any insights, multiple tabs with no clear starting point, and complete cognitive overload
- This is precisely why dashboards get abandoned - nobody wants this experience
- To solve these problems, we developed a systematic approach that addresses both user experience and the psychological aspects of data interpretation
:::

@@

- Our framework flows through 5 stages: Notice → Interpret → Structure → Anticipate → Validate
  * Click through
- Notice identifies friction points where users struggle with the interface
- Interpret uncovers what story users truly need - moving from symptoms to diagnosis
- Structure organizes information in ways that align with how our brains process information
- Anticipate considers cognitive biases affecting data interpretation
- Validate ensures users leave with clear takeaways and a positive impression
- Remember, these stages are iterative - you'll often revisit earlier stages as you discover new insights
:::

@@

- BID integrates established best practices from multiple disciplines - we're truly standing on the shoulders of giants
  * Click through example fields...UX, data storytelling, psych science, data/visual communication
- What makes BID uniquely powerful is how it systematically connects psychological principles to specific coding decisions at each development stage
- Unlike typical checklists, BID guides how you think about dashboards from psychological first principles
- The framework bridges theory and practice - showing when and how to apply principles in your workflow
- You don't need a psychology degree to apply these effectively
- Let's see how these stages transformed SoundPulse's dashboard
:::

@@

- **Stage 1 (Notice):** Users were overwhelmed with options
  * This relates to 'cognitive load' - the mental effort needed to process information
  * When the brain gets overtaxed, comprehension drops
  * Hick's Law shows decision time increases logarithmically with more choices
- **Stage 2 (Interpret):** We identified what users needed - increased Processing Fluency
  * When information is easy to process, people find it more credible
  * We focused on the key story executives needed, not showing all possible data
- **Stage 3 (Structure):** Moving to the Structuring stage...applied the 'Principle of Proximity' - which is a fancy way of saying we started grouping related elements together. 
  * Dual-Processing Theory distinguishes between fast, intuitive thinking and slower, deliberative thinking - we designed for both
  * We created logical groupings of metrics for natural data flow
- Now let's look at the code implementation
:::

@@

- For the later BID stages, we addressed subtler psychological factors affecting interpretation and action
- **Stage 4 (Anticipate):** We expected some folks to make flawed comparisons if they didn't have proper context.
  * This relates to the 'Anchoring Effect' - where people rely too heavily on the first piece of information they see
  * We also saw teams struggling to interpret the data in the same way, so we addressed this with something called a "framing toggle"...more on that in just a moment
- **Stage 5 (Validate):** Finally, we applied the 'Peak-End Rule' - the idea that people judge experiences by their most intense moment and ending
  * We created summary cards highlighting key takeaways so users leave with a clear, memorable impression
:::

@@

- This illustrates the powerful psychological principles at work...we're showing identical data but creating completely different mental states
- Progress framing highlights achievement and builds confidence...marketing team preferred this approach
- Gap framing creates urgency and focuses on improvements needed - product teams used this view
- "This isn't just theory - a classic study showed that people react completely differently when presenting identical medical treatment outcomes as either a 90% survival rate or a 10% mortality rate ... even though it's the same data"
- Consider how framing in your dashboards might drive different decisions
:::

@@

- We're not just sharing theories - we're giving you practical tools for immediate implementation
- The `{bidux}` package directly implements our 5-stage framework
- Each function maps to a specific stage and guides your thinking
- Phase 1 is available today on GitHub with the conceptual foundation
- Future phases will include LLM integration for suggesting implementations
- We're also developing pre-built UI components embodying these principles
- Think of `{bidux}` as your behavioral science assistant for Shiny development
:::

@@

- "Let's look at what {bidux} offers today"
- "The concept dictionary provides implementation advice, not just theory"
- "You can search across psychology, UI/UX, and data visualization fields...and more!"
- "The workflow functions guide you through each BID stage"
- "Auto-suggestion of theories based on your problem description"
- "Component suggestions provide concrete UI implementation ideas"
- When the link appears: "We're launching this at ShinyConf today - check out the repo for full documentation"
:::

@@

- The results speak for themselves
- While quantitative metrics are still being collected, qualitative feedback has been overwhelmingly positive
- Users who previously abandoned dashboards now rely on them daily
- Decisions that took days now happen in minutes
- When testimonial appears: The emerging message is clear: this framework transforms how teams interact with their data
- "Remember, this wasn't just about visual improvements - this was a fundamental shift in how our users...and developers...approached the dashboard"
:::

@@

- We'd love to hear about your dashboard challenges"
- Connect with us on LinkedIn/GitHub for updates on {bidux} development
- Feel free to reach out for questions or deeper discussions
- Quick answers to likely questions:
  * "How do you measure success?" → "Start with user testing and session analytics; we're building telemetry tools to help quantify impact"
  * "Does this work for technical audiences?" → "Absolutely, technical users appreciate clarity too - in fact, they often have the highest cognitive load due to the complexity of their work"
  * "How long does implementation take?" → "Start small - even one principle makes a difference. A complete BID implementation can take 1-2 sprints depending on dashboard complexity"
  * "How does this apply to domains beyond music?" → "We've successfully applied these principles in healthcare, finance, and retail dashboards - the psychological principles are universal"
:::
